Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-02-23 paper/main.tex
intro and part of prelim. polished

\title{Reducing Acceptance Marks\\ in Emerson-Lei Automata by QBF-solving}
\author{Tereza Schwarzov\'{a}}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
\funding{This work has been supported by the Czech Science Foundation grant GA19-24397S.}

@@

\begin{abstract}
  \emph{Transition-based Emerson-Lei automata (TELA)} become a popular
  formalism that generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are labelled with acceptance marks
  and its accepting formula is a positive Boolean combination of terms
  saying that a particular mark has to be visited infinitely or
  finitely often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique reducing the number
  of %different
  acceptance marks in TELA with use of \emph{quantified Boolean
    formulas (QBF)}: if a constructed QBF formula is satisfiable then
  its model encodes an equivalent accepting formula with less
  acceptance marks. In fact, we present three formula constructions
  trading the speed of the technique for higher reduction
  efficiency. We evaluated our reduction technique on TELA produced by
  state-of-the-art tools Delag, ltl3tela, Rabinizer~4, and SPOT. The
  technique reduced some acceptance marks in automata produced by all
  these tools. In the case of Rabinizer~4, the cumulative number of
  acceptance marks was reduced to less than one half.
\end{abstract}

@@

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity automata~\cite{renkin.20.atva}. \todo{Doplnit!}

@@

\todo{Honza je tu!}

These observations motivate our study of techniques reducing the
number of acceptance marks without altering the structure of the
automaton. We present basically two reduction techniques. The first
one is based on relations among individual transition sets defined by
various acceptance marks in each \emph{strongly connected component
  (SCC)}. This technique can be seen as an extension of the
\emph{SCC-base simplifications of acceptance
  conditions}~\cite{babiak.13.spin} designed for
\emph{transition-based generalized Büchi automata (TGBA)} and
implemented in SPOT. Probably the main difference between the original
technique for TGBA and the one for TELA is in the final step when
simplified acceptance conditions for each SCC are merged together into
one reduced acceptance condition.

The second technique is completely different. For a given TELA and
parameters $c,m$, it produces a \emph{quantified Boolean formula
  (QBF)} which is satisfiable if and only if there exists an automaton
with the same structure, $m$ acceptance marks, an acceptance formula
in disjunctive normal form with $c$ clauses, and the same set
of accepting runs. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of a
simpler formula such its satisfiability implies existence of an
automaton with the same structure, $m$ acceptance marks, and the same
set of accepting runs, but not vice versa.

We have implemented both reduction techniques in a tool called
\texttt{telatko}. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav},
and SPOT~\cite{duret.16.atva2}.  While the reduction is relatively
modest for TELA produced by ltl3tela and SPOT, it is substantial for
automata produced by Rabinizer~4.

\subsubsection{Related results}
We are not aware about any existing research aimed at simplification
of acceptance formulas of TELA or reduction of the number of its
acceptance marks without increasing the number of states.

There are some SAT-based approaches aimed to reduce the number of
states of automata over infinite words. Namely, there are reductions
designed for nondeterministic Büchi automata~\cite{ehlers.10.spin},
deterministic Büchi automata~\cite{ehlers.10.sat}, and deterministic
generalized Büchi automata~\cite{baarir.14.forte}. Further, there is
also a SAT-based approach that transforms a deterministic TELA
automaton to an equivalent automaton with a given acceptance condition
and a given number of states~\cite{baarir.15.lpar} (if such an
automaton exists). Note that all these techniques are usually very
slow and their authors typically suggest to use them only for specific
purposes like looking for cases where some automata construction can
be improved.

Casares, Colcombet, and Fijalkow very recently introduced a structure
called \emph{alternating cycle decomposition
  (ACD)}~\cite{casares.21.icalp} which compactly represents the
information about all accepting and non-accepting automata cycles.  We
expect that ACD could potentialy be used to reduce the number of
acceptance marks or to simplify the acceptance condition. However,
such a reduction is not obvious. Moreover, the construction of ACD
seems to be computationally expensive.

\subsubsection{Structure of the paper} The next section introduces
basic terms used in the paper. Section~\ref{sec:standard} presents the
reduction technique based on relations among individual acceptance
marks on each SCC. The QBF-based technique is explained in
Section~\ref{sec:qbf}. Implementation is briefly described in
Section~\ref{sec:experiments}, which also shows the experimental
results. Finally, Section~\ref{sec:conclusion} suggests other
applications of our QBF-reduction technique and closes the paper.

@@

%A transition $(p,a,M',q)\in\delta$ is a transition leading from state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$
A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $M(\pi)$ denote set of all acceptance marks that occur
in run $\pi$ in infinitely many transitions. Run $\pi$ is
\emph{accepting} iff it satisfies the formula $\varphi$, where $\pi$
satisfies $\Inf(m)$ iff $m\in M(\pi)$ and $\pi$ satisfies $\Fin(m)$
iff $m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the
set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

%}}}
%{{{ standard reductions

@@

%}}}
%{{{ implementation and experimental evaluation

\section{Implementation and experimental evaluation}\label{sec:experiments}

@@

>>>>

\title{Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving}
\author{Tereza Schwarzov\'{a}}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
\funding{This work has been supported by the Czech Science Foundation grant GA19-24397S.}

@@

\begin{abstract}
  \emph{Transition-based Emerson-Lei automata (TELA)} become a popular
  formalism that generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are labelled with acceptance marks
  and its accepting formula is a positive Boolean combination of terms
  saying that a particular mark has to be visited infinitely or
  finitely often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique reducing the number
  of %different
  acceptance marks in TELA with use of \emph{quantified Boolean
    formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  In fact, we present three formula constructions
  trading speed for higher reduction
  efficiency. We evaluated our reduction technique on TELA produced by
  state-of-the-art tools Delag, ltl3tela, Rabinizer~4, and Spot. The
  technique reduced some acceptance marks in automata produced by all
  these tools. In the case of Rabinizer~4, the cumulative number of
  acceptance marks was reduced to less than one half.
\end{abstract}

@@

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

These observations motivate our study of a technique reducing the
number of acceptance marks without altering the structure of the
automaton. This paper presents a technique heavily based on
\emph{quantified Boolean formulas (QBF)}. For a given TELA and
parameters $c,m$, it produces a QBF which is satisfiable if and only
if there exists an automaton with the same structure, $m$ acceptance
marks, an acceptance formula in disjunctive normal form with $c$
clauses, and the same set of accepting runs as the original
automaton. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of
two simpler formulas such that their satisfiability implies the
existence of an automaton with the same structure, $m$ acceptance
marks, and the same set of accepting runs, but not vice versa.

We have implemented our reduction technique in a tool called
\texttt{telatko}. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav},
and Spot~\cite{duret.16.atva2}.  While the reduction is relatively
modest for TELA produced by ltl3tela and Spot, it is substantial for
automata produced by Rabinizer~4. \todo{A jak vyjde Delag? Fixnout i
  Abstract.}

\paragraph*{Related results}
We are not aware about any existing research primarily aimed at
simplification of acceptance formulas of TELA or reduction of the
number of its acceptance marks without increasing the number of
states. There exists only a SAT-based approach that transforms a
deterministic TELA automaton to an equivalent automaton with a given
acceptance condition and a given number of
states~\cite{baarir.15.lpar} (if such an automaton exists). Further,
there are some SAT-based approaches aimed to reduce the number of
states of automata over infinite words. More preceisely, there are
reductions designed for nondeterministic Büchi
automata~\cite{ehlers.10.spin}, deterministic Büchi
automata~\cite{ehlers.10.sat}, and deterministic generalized Büchi
automata~\cite{baarir.14.forte}. Note that all these techniques are
usually very slow and their authors typically suggest to use them only
for specific purposes like looking for cases where some automata
construction can be improved.

Casares, Colcombet, and Fijalkow very recently introduced a structure
called \emph{alternating cycle decomposition
  (ACD)}~\cite{casares.21.icalp} which compactly represents the
information about all accepting and non-accepting automata cycles. We
expect that ACD could potentialy be used to reduce the number of
acceptance marks or to simplify the acceptance condition. However,
such a reduction is not obvious.
%Moreover, the construction of ACD seems to be computationally expensive.

\paragraph*{Structure of the paper} The next section introduces basic
terms used in the paper. Section~\ref{sec:qbf} explains the
construction of three mentioned QBF formulas. The reduction algorithm
based on these formulas is described together with its implementation
in Section~\ref{sec:algimpl}. Section~\ref{sec:experiments} shows the
experimental results. Finally, Section~\ref{sec:conclusion} suggests
other applications of our QBF-reduction technique and closes the
paper.

@@

%A transition $(p,a,M',q)\in\delta$ is a transition leading from state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$
A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $M(\pi)$ denote set of all acceptance marks that occur
in infinitely many transitions of run $\pi$. Run $\pi$ is
\emph{accepting} iff it satisfies the formula $\varphi$, where $\pi$
satisfies $\Inf(m)$ iff $m\in M(\pi)$ and $\pi$ satisfies $\Fin(m)$
iff $m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the
set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

\todo{Jan is here}

@@

@@

%}}}
%{{{ algorithm and implementation

\section{Reduction algorithm and its implementation}\label{sec:algimpl}
Jak se nastavuji inicialni parametry a jak postupne klesaji.
Jak prechazim od nizsich levelu k vyssim.
Vyuzivame incremental QBF solving?

%}}}
%{{{ experimental evaluation

\section{Experimental evaluation}\label{sec:experiments}

@@

%{{{ standard reductions
